Theory of Phonon Polaritons in Polar Dielectrics

### 2018-08-06 paper.tex
eq and fig refs and addedCSGF fellow number (jenny i think yours might be different)

\usepackage[usenames]{color}
\newcommand{\edited}[1]{{\color{red} #1}}

@@

We start by developing the theory of phonon polaritons in a 3D bulk in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties. Here, the matter subsystem which couples to light is the ionic lattice of the crystal. Thus, we develop the linear response theory of an ionic lattice coupled to an electromagnetic field. The interaction Hamiltonian of the lattice with the electromagnetic field is taken as:
\begin{equation}
H_{int} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
\end{equation}
where $\mathbf{E}$ is a driving electric field, and $\mathbf{P}$ is the polarization associated with the lattice. We write the polarization as a sum of contributions from all unit cells in the system. Namely
\begin{equation}
\mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
\end{equation}
where $\mathbf{d}(\mathbf{R},t)$ is the time-dependent dipole moment associated with lattice site $\mathbf{R}$. In expressing the polarization in this fashion, we are assuming that the displacement of the ions is far smaller than the size of the unit cell, which is, even in a deeply anharmonic regime, a good approximation. This said, we operate in the harmonic regime.  The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa} \mathbf{b}_{\kappa} \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$. Note that the Born effective charges are both tensorial and site-dependent.  This displacement can be expanded in phonon modes as
\begin{equation}
\mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
\end{equation}
where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation (1). We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.

To connect this polarization-polarization response function to the observed dielectric function, we write down Maxwell's equations for the electric field sourced by a polarization density. We have that
\begin{equation}
\left(-\mathbf{q}\times\mathbf{q}\times - \epsilon_{\infty}\frac{\omega^2}{c^2} \right)\mathbf{E} = \omega^2\mu_0\mathbf{P}(\mathbf{q},\omega),
\end{equation}
where $\omega$ is the frequency of light under consideration $\mathbf{P}$ is the polarization at that frequency, and $\epsilon_{\infty}$ is the high-frequency dielectric function associated with the electrons. Now, we identify the average polarization $\langle\mathbf{P}\rangle$ with the classical polarization $\mathbf{P}$ in Maxwell's equations. This is essentially the random-phase approximation (RPA). We note that an analogous procedure is also employed in the linear-response theory of the longitudinal response of electrons, where instead of the polarization density being averaged, the charge density is averaged. And instead of the solving Maxwell's equations for the field, we solve Laplace's equation for the potential. Using the linear response relation $\mathbf{P} = \epsilon_0\boldsymbol{\Pi}\mathbf{E}$, we have simply that
\begin{equation}
\left(-\mathbf{q}\times\mathbf{q}\times - \left(\epsilon_{\infty} + \boldsymbol{\Pi} \right)\frac{\omega^2}{c^2}\right)\mathbf{E} = 0,
\end{equation}
leading to the identification of of the dielectric tensor $\boldsymbol{\epsilon} = \epsilon_{\infty} + \boldsymbol{\Pi}$.

@@

\begin{equation}

@@

\begin{equation}

@@

where it is to be understood that $\mathbf{P}(\mathbf{q})$ refers to a continuous Fourier transform while $\mathbf{d}(\mathbf{q})$ refers to a discrete Fourier transform. Substituting in the displacement operator of Equation (3), we find that $\mathbf{d}(\mathbf{q})$ is given by
\begin{equation}

@@

\begin{equation}

@@

This can be expressed in terms of more standard outputs of \emph{ab initio} materials physics methods that calculate phonon properties. In particular, we can take $\hat{e}_{\mathbf{q}\sigma}$, the eigenvectors of the dynamical matrix, and define eigendisplacements via $\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}=\frac{\hat{e}_{\mathbf{q}\sigma}}{\sqrt{M_{\kappa}}}$. In which case, we may cast Equation (9) into the form:
\begin{equation}

@@

We may now use this in concurrence with Equation (6) to evaluate the dielectric function. Consider $\boldsymbol{\Pi}(\mathbf{q},\omega)$. From the form of Equation (8), we see that at zero temperature, the first term only sees contributions from $m=|0_{\mathbf{q}\sigma}\rangle$ and $n=|1_{\mathbf{q}\sigma}\rangle$, where $0_{\mathbf{q}\sigma}$ and $1_{\mathbf{q}\sigma}$ refer to the number of phonons in mode $\mathbf{q}\sigma$. Here $\sigma$ can be any branch of the phonon band structure. In the second term, $m$ and $n$ switch. In the limit where the phonons are very weakly damped, $\Gamma_{\mathbf{q}\sigma} \\ \omega_{\mathbf{q}\sigma}$. this leads to:
\begin{equation}

@@

\begin{equation}

@@

\begin{equation}

@@

Before applying this formalism to a real material and discussion of the two-dimensional case, we briefly consider and comment on `nonlocality' in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.  Therefore, it is possible that the phonon polariton resonances red-shift or blue-shift, depending on how the relative position of the phonons at finite-wavevector versus zero wavevector. For example, in 4H-SiC, upon moving along the $\Gamma-L$ direction of the Brillouin zone, the optical phonon which determines the phonon polariton, calculated to be at 764 cm$^{-1}$ at $\Gamma$, red-shifts by 2  cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, by 5 cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, and 13 cm$^{-1}$ at 1/2 of the way along the $\Gamma-L$ direction. This will lead to a red-shift of the Reststrahlen bands by these amounts.
\begin{figure*}[t]
\includegraphics[width=16cm]{combined_structure.pdf}
\caption{\textbf{Crystal structure and phonon dispersion of bulk 4H silicon carbide and hexagonal boron nitride.} (a) The hexagonal unit cell structure of silicon carbide allows different optical response in the a-b plane and along the c axis. (b) The full phonon bandstructure of silicon carbide, from which the phonon eigendisplacements associated with the gamma point were used to calculate the frequency-dependent dielectric function. Structure (looking from stacking plane) (c) and phonon bandstructure (d) as before but for bulk hexagonal boron nitride.}
\label{fig:phonons}
\end{figure*}

@@

%This is exemplified in Figure 3, where we show the ionic contribution to the permittivity (i.e. we do not consider the electronic part). We consider wavevector directions along $\Gamma-L$ and consider wavevectors that are $1/10$ and $1/4$ of the way along the Brillouin zone. We see that the major effect is to red-shift the phonon resonance by a few inverse centimeters. However, the effect "downstream", in the region where phonon polaritons are, is much smaller, with the permittivity being almost the same as in the local case.  In making this plot, we make the assumption that the variation of the Born effective charges is negligible upon moving to finite-wavevector optical phonons.

%However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.

%section below appears twice - commenting it out.

%Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on `nonlocality' in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion. As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components. However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$. In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then appear  to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.

\section{Application to bulk semiconductor systems}
% calculation basics
To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a). 
Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting. Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}. We calculated the phonon modes, their frequencies (shown in Figure 1(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation (12), using experimental values for the dissipation rates $\Gamma$. We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.

% about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
Additionally, for SiC a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}. This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response. In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}. 
Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent \emph{ab initio} calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV. As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
From these calculations, the high-frequency dielectric tensor for SiC was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986, $\epsilon^{\infty}_{\perp}$ = 7.307 and for hBN, $\epsilon^{\infty}_{\parallel}$ = 4.634, $\epsilon^{\infty}_{\perp}$ = 2.593.  This also leads to a modification of the phonon spectra.
Note that for a materials where the optical band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used while retaining the \emph{ab initio} implementation of our theory. Further, if excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter Equation prediction might be used to further improve predictive accuracy within the same framework presented here.
The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}. As can be seen, the agreement between the theoretical calculation and the experimental results is quite good for silicon carbide, and good for the upper Reststrahlen band of hBN. For the lower Reststrahlen band, the major discrepancy results from a roughly 5\% deviation between calculated and measured transverse optical phonon frequencies.

% can also put this table just by listing eps^inf_perp and parallel in the main text.  
\begin{figure}[t]
\includegraphics[width=8.5cm]{SiC_comparison_thy_vs_expt.pdf}
\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk 4H silicon carbide.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk 4H silicon carbide from using Equation (12). The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package. Overlaid in blue dots are the experimental values for the dielectric function, taken from \cite{tiwald1999carrier}. The real part is shown in (a) while the imaginary part is shown in (b). }
\label{fig:epsilon}
\end{figure}

\begin{figure}[t]
\includegraphics[width=8.5cm]{hBN_comparison_thy_vs_expt.pdf}
\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk hexagonal boron nitride.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk hexagonal boron nitride from using Equation (12). The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package. Overlaid in blue dots are the experimental values for the dielectric function, taken from \cite{caldwell2014sub}. The component of the real part of the permittivity parallel to the c-axis is shown in (a) component parallel is shown in (b).}
\label{fig:phonons}
\end{figure}

@@

For $z \neq 0$, the Maxwell equation for the magnetic field is $(\nabla\times\nabla\times - \epsilon\frac{\omega^2}{c^2})\mathbf{H} = 0$. This equation, for the TM mode described above, becomes 
\begin{equation}
\left(-\frac{d^2}{dz^2}+\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2} \right)H = 0,
\end{equation}
where we have written $\mathbf{H} = H\hat{y}$. The solution in the region $z > 0$ is simply \begin{align}
H &= e^{i\mathbf{q}\cdot\mathbf{r}_{||}-\kappa z}, ~z>0 \nonumber \\
H &= -e^{i\mathbf{q}\cdot\mathbf{r}_{||}-\kappa z}, ~z>0 \nonumber \\
\end{align}
where $\kappa = \sqrt{\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2}}$ and we have implicitly assumed $q > \epsilon\frac{\omega}{c}$ as we are interested in confined evanescent modes. To join these two solutions consistently, we apply the interface conditions for the magnetic field. They are $\hat{z}\times(\mathbf{H}^{(+)}-\mathbf{H}^{(-)}) = \mathbf{K} = -i\omega\mathbf{P}_s$ where $\mathbf{K}$ is the surface current density, expressed through the surface polarization density $\mathbf{P}_s$. 
Now we use the linear response relation $\mathbf{P}(\mathbf{q},\omega) = \boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$ in addition to Ampere's law $\nabla\times\mathbf{H} = -i\omega\epsilon\mathbf{E}$. Plugging these relations in yields an implicit equation connecting the polarization response function, the frequency of the 2D phonon polariton, and its frequency. Taking the case where $x$ coincides with a principal axis of the system, we have that 
\begin{equation}
q\Pi_{xx}(q,\omega)  = -2\epsilon.
\end{equation}

>>>>

%\usepackage[usenames]{color}
%\newcommand{\edited}[1]{{\color{red} #1}}

@@

We start by developing the theory of phonon polaritons in a 3D bulk in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties. Here, the matter subsystem which couples to light is the ionic lattice of the crystal. Thus, we develop the linear response theory of an ionic lattice coupled to an electromagnetic field. The interaction Hamiltonian of the lattice with the electromagnetic field is taken as:
\begin{equation}\label{eq:interaction}
H_{\text{int}} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
\end{equation}
where $\mathbf{E}$ is a driving electric field, and $\mathbf{P}$ is the polarization associated with the lattice. We write the polarization as a sum of contributions from all unit cells in the system. Namely
\begin{equation}\label{eq:dipolesum}
\mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
\end{equation}
where $\mathbf{d}(\mathbf{R},t)$ is the time-dependent dipole moment associated with lattice site $\mathbf{R}$. In expressing the polarization in this fashion, we are assuming that the displacement of the ions is far smaller than the size of the unit cell, which is, even in a deeply anharmonic regime, a good approximation. This said, we operate in the harmonic regime.  The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa}  \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$. Note that the Born effective charges are both tensorial and site-dependent.  This displacement can be expanded in phonon modes as
\begin{equation}\label{eq:phononfield}
\mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
\end{equation}
where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation \ref{eq:interaction}. We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.

To connect this polarization-polarization response function to the observed dielectric function, we write down Maxwell's equations for the electric field sourced by a polarization density. We have that
\begin{equation}\label{eq:maxwellsourced}
\left(-\mathbf{q}\times\mathbf{q}\times - \epsilon_{\infty}\frac{\omega^2}{c^2} \right)\mathbf{E} = \omega^2\mu_0\mathbf{P}(\mathbf{q},\omega),
\end{equation}
where $\omega$ is the frequency of light under consideration $\mathbf{P}$ is the polarization at that frequency, and $\epsilon_{\infty}$ is the high-frequency dielectric function associated with the electrons. Now, we identify the average polarization $\langle\mathbf{P}\rangle$ with the classical polarization $\mathbf{P}$ in Maxwell's equations. This is essentially the random-phase approximation (RPA). We note that an analogous procedure is also employed in the linear-response theory of the longitudinal response of electrons, where instead of the polarization density being averaged, the charge density is averaged. And instead of the solving Maxwell's equations for the field, we solve Laplace's equation for the potential. Using the linear response relation $\mathbf{P} = \epsilon_0\boldsymbol{\Pi}\mathbf{E}$, we have simply that
\begin{equation}\label{eq:maxwellepsilon}
\left(-\mathbf{q}\times\mathbf{q}\times - \left(\epsilon_{\infty} + \boldsymbol{\Pi} \right)\frac{\omega^2}{c^2}\right)\mathbf{E} = 0,
\end{equation}
leading to the identification of of the dielectric tensor $\boldsymbol{\epsilon} = \epsilon_{\infty} + \boldsymbol{\Pi}$.

@@

\begin{equation}\label{eq:polarizationkubo}

@@

\begin{equation}\label{eq:fourierpolarization}

@@

where it is to be understood that $\mathbf{P}(\mathbf{q})$ refers to a continuous Fourier transform while $\mathbf{d}(\mathbf{q})$ refers to a discrete Fourier transform. Substituting in the displacement operator of Equation \ref{eq:phononfield}, we find that $\mathbf{d}(\mathbf{q})$ is given by
\begin{equation}\label{eq:phononstodipole}

@@

\begin{equation}\label{eq:oscillatordefinition}

@@

This can be expressed in terms of more standard outputs of \emph{ab initio} materials physics methods that calculate phonon properties. In particular, we can take $\hat{e}_{\mathbf{q}\sigma}$, the eigenvectors of the dynamical matrix, and define eigendisplacements via $\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}=\frac{\hat{e}_{\mathbf{q}\sigma}}{\sqrt{M_{\kappa}}}$. In which case, we may cast Equation \ref{eq:oscillatordefinition} into the form:
\begin{equation}\label{eq:oscillatorstrength}

@@

We may now use this in concurrence with Equation \cite{eq:polarizationkubo} to evaluate the dielectric function. Consider $\boldsymbol{\Pi}(\mathbf{q},\omega)$. From the form of Equation \ref{eq:phononstodipole}, we see that at zero temperature, the first term only sees contributions from $m=|0_{\mathbf{q}\sigma}\rangle$ and $n=|1_{\mathbf{q}\sigma}\rangle$, where $0_{\mathbf{q}\sigma}$ and $1_{\mathbf{q}\sigma}$ refer to the number of phonons in mode $\mathbf{q}\sigma$. Here $\sigma$ can be any branch of the phonon band structure. In the second term, $m$ and $n$ switch. In the limit where the phonons are very weakly damped, $\Gamma_{\mathbf{q}\sigma} \\ \omega_{\mathbf{q}\sigma}$. this leads to:
\begin{equation}\label{eq:fullsusceptibility}

@@

\begin{equation}\label{eq:pitoepsilon}

@@

\begin{equation}\label{eq:lorentzoscillator}

@@

Before applying this formalism to a real material and discussion of the two-dimensional case, we briefly consider and comment on `nonlocality' in the phonon contribution to the dielectric function. Note that Equation \ref{eq:pitoepsilon} has a wavevector dependence which is implicit through the phonon dispersion.  Therefore, it is possible that the phonon polariton resonances red-shift or blue-shift, depending on how the relative position of the phonons at finite-wavevector versus zero wavevector. For example, in 4H-SiC, upon moving along the $\Gamma-L$ direction of the Brillouin zone, the optical phonon which determines the phonon polariton, calculated to be at 764 cm$^{-1}$ at $\Gamma$, red-shifts by 2  cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, by 5 cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, and 13 cm$^{-1}$ at 1/2 of the way along the $\Gamma-L$ direction. This will lead to a red-shift of the Reststrahlen bands by these amounts.
\begin{figure*}[t]
\includegraphics[width=16cm]{combined_structure.pdf}
\caption{\textbf{Crystal structure and phonon dispersion of bulk 4H silicon carbide and hexagonal boron nitride.} (a) The hexagonal unit cell structure of silicon carbide allows different optical response in the a-b plane and along the c axis. (b) The full phonon bandstructure of silicon carbide, from which the phonon eigendisplacements associated with the gamma point were used to calculate the frequency-dependent dielectric function. Structure (looking from stacking plane) (c) and phonon bandstructure (d) as before but for bulk hexagonal boron nitride.}
\label{fig:phonons}
\end{figure*}

@@

\section{Application to bulk semiconductor systems}
% calculation basics
To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in \ref{fig:phonons}(a). 
Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting. Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}. We calculated the phonon modes, their frequencies (shown in Figure \ref{fig:phonons}(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation \ref{eq:pitoepsilon}, using experimental values for the dissipation rates $\Gamma$. We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.

% about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
Additionally, for SiC a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}. This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response. In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}. 
Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent \emph{ab initio} calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV. As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
From these calculations, the high-frequency dielectric tensor for SiC was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986, $\epsilon^{\infty}_{\perp}$ = 7.307 and for hBN, $\epsilon^{\infty}_{\parallel}$ = 4.634, $\epsilon^{\infty}_{\perp}$ = 2.593.  This also leads to a modification of the phonon spectra.
Note that for a materials where the optical band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used while retaining the \emph{ab initio} implementation of our theory. Further, if excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter Equation prediction might be used to further improve predictive accuracy within the same framework presented here.
The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figures \ref{fig:sicepsilon} and \ref{fig:hbnepsilon}. As can be seen, the agreement between the theoretical calculation and the experimental results is quite good for silicon carbide, and good for the upper Reststrahlen band of hBN. For the lower Reststrahlen band, the major discrepancy results from a roughly 5\% deviation between calculated and measured transverse optical phonon frequencies.

% can also put this table just by listing eps^inf_perp and parallel in the main text.  
\begin{figure}[t]
\includegraphics[width=8.5cm]{SiC_comparison_thy_vs_expt.pdf}
\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk 4H silicon carbide.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk 4H silicon carbide from using Equation \ref{eq:pitoepsilon}. The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package. Overlaid in blue dots are the experimental values for the dielectric function, taken from \cite{tiwald1999carrier}. The real part is shown in (a) while the imaginary part is shown in (b). }
\label{fig:sicepsilon}
\end{figure}

\begin{figure}[t]
\includegraphics[width=8.5cm]{hBN_comparison_thy_vs_expt.pdf}
\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk hexagonal boron nitride.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk hexagonal boron nitride from using Equation \ref{eq:pitoepsilon}. The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package. Overlaid in blue dots are the experimental values for the dielectric function, taken from \cite{caldwell2014sub}. The component of the real part of the permittivity parallel to the c-axis is shown in (a) component parallel is shown in (b).}
\label{fig:hbnepsilon}
\end{figure}

@@

For $z \neq 0$, the Maxwell equation for the magnetic field is $(\nabla\times\nabla\times - \epsilon\frac{\omega^2}{c^2})\mathbf{H} = 0$. This equation, for the TM mode described above, becomes 
\begin{equation}\label{eq:2dmaxwell}
\left(-\frac{d^2}{dz^2}+\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2} \right)H = 0,
\end{equation}
where we have written $\mathbf{H} = H\hat{y}$. The solution in the region $z > 0$ is simply \begin{align}
H &= e^{i\mathbf{q}\cdot\mathbf{r}_{||}-\kappa z}, ~z>0 \nonumber \\
H &= -e^{i\mathbf{q}\cdot\mathbf{r}_{||}-\kappa z}, ~z>0 \nonumber \\
\end{align}
where $\kappa = \sqrt{\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2}}$ and we have implicitly assumed $q > \epsilon\frac{\omega}{c}$ as we are interested in confined evanescent modes. To join these two solutions consistently, we apply the interface conditions for the magnetic field. They are $\hat{z}\times(\mathbf{H}^{(+)}-\mathbf{H}^{(-)}) = \mathbf{K} = -i\omega\mathbf{P}_s$ where $\mathbf{K}$ is the surface current density, expressed through the surface polarization density $\mathbf{P}_s$. 
Now we use the linear response relation $\mathbf{P}(\mathbf{q},\omega) = \boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$ in addition to Ampere's law $\nabla\times\mathbf{H} = -i\omega\epsilon\mathbf{E}$. Plugging these relations in yields an implicit equation connecting the polarization response function, the frequency of the 2D phonon polariton, and its frequency. Taking the case where $x$ coincides with a principal axis of the system, we have that 
\begin{equation}\label{eq:2ddispersion}
q\Pi_{xx}(q,\omega)  = -2\epsilon.
\end{equation}

To connect this relation to the discussion of the three-dimensional case, we write a microscopic form for the polarization response function of the ionic lattice. It is simply
\begin{equation}\label{eq:2dsusceptibility}
\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 A}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i0^+}\left(e^{\beta E_m}-e^{\beta E_n} \right),
\end{equation}
where the only change is to replace a normalization volume with a normalization area $A$. Here, $\mathbf{P}(\mathbf{q})$ is the 2D Fourier transform of the polarization density. As the form of the displacement of the ions is still given by Equation (3), $\mathbf{P}(\mathbf{q})$ is still given by Equations (7-10), but with the Born charges and eigendisplacements appropriate to the 2D system of interest. Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix. Following the same kind of reasoning that lead to Equation \ref{eq:lorentzoscillator}, if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation \ref{eq:lorentzoscillator} but with $V$ replaced by $A$. Defining $N/A \equiv n_s$, Equation \ref{eq:2dmaxwell} for the dispersion relation of 2D phonon polaritons tells us that
\begin{equation}\label{eq:2dphononpolaritondispersion}
\omega = \sqrt{\omega_{TO}^2+\frac{n_sQ^{*2}}{2M^*\epsilon_0\epsilon}q}.
\end{equation}
Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form\cite{stern1967polarizability} $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ . The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency.
